---
title: "Lab 07 - Conveying the right message through visualisation"
author: "Tsion"
output: github_document
---

### Load packages and data

```{r load}
library(tidyverse)
library(scales)
library(lubridate)

# library(zoo) didn't work for me so I tried installing it in the console instead

df <- read_csv("kansas_grouped_rolling_avg.csv")
```

## Part 1: The Problem — Where's the Data?

### Exercise 1

I think the original visualization appears to use data from **Kansas** during the summer of 2020. The title says it is showing the **Kansas COVID-19 7-day rolling average of daily cases per 100K population**. 

It compares two groups: **counties with mask mandates** and **counties without mask mandates**. I believe the phrase “per 100K population” means the case counts were adjusted for population size, so counties or groups with different populations can be compared more fairly.


### Exercise 2

I think the MMWR report gives two important details that were missing or not obvious in the original visualization.

It names the 24 Kansas counties that had mask mandates in effect by July 3, 2020. These were:

Allen, Atchison, Bourbon, Crawford, Dickinson, Douglas, Franklin, Geary, Gove, Harvey, Jewell, Johnson, Mitchell, Montgomery, Morris, Pratt, Reno, Republic, Saline, Scott, Sedgwick, Shawnee, Stanton, and Wyandotte.

It also says the COVID-19 case data came from USAFacts, which provides county-level COVID-19 data. This is important because it lets us download the raw county-level case data and reconstruct the visualization ourselves.

### Part 2: Acquiring the Raw Data

### Exercise #3

```{r Download the confirmed cases data and the county population data directly from USAFacts}
library(tidyverse)
library(scales)
library(lubridate)
cases_raw <- read_csv(
  "https://static.usafacts.org/public/data/covid-19/covid_confirmed_usafacts.csv",
  show_col_types = FALSE
)

pop_raw <- read_csv(
  "https://static.usafacts.org/public/data/covid-19/covid_county_population_usafacts.csv",
  show_col_types = FALSE
)
```

```{r explore both datasets}
glimpse(cases_raw)
glimpse(pop_raw)
```

That was a lot of text wheww! The cases data has `r nrow(cases_raw)` rows and `r ncol(cases_raw)` columns. Each row represents one county or county-equivalent area in the United States, with identifying information like county name, state, and FIPS code, followed by cumulative COVID-19 case counts for many dates.

The data is in wide format because each date is stored as its own column. As we learned before, I think this is a problem for plotting with `ggplot2` because it works best when the variable that changes over time is stored in one column. For a time series graph, it is better to have one column for `date` and one column for `cases`, so we need to reshape the data into long format before plotting (knowing this also helped me a lot while doing my plots for my eye tracking study as well :) ).

### Exercise 4

```{r filter-kansas-and-pivot-longer}
# Keep only Kansas county rows and remove the statewide total row
ks_cases <- cases_raw %>%
  filter(State == "KS", countyFIPS != 0) %>%
  pivot_longer(
    cols = matches("^\\d{4}-\\d{2}-\\d{2}$"),
    names_to = "date",
    values_to = "cumulative_cases"
  ) %>%
  mutate(date = ymd(date)) %>%
  select(countyFIPS, `County Name`, date, cumulative_cases)

# Keep only Kansas county population data
ks_pop <- pop_raw %>%
  filter(State == "KS", countyFIPS != 0) %>%
  select(countyFIPS, `County Name`, population)

# Check results
glimpse(ks_cases)
glimpse(ks_pop)
```

## Part 3: From Cumulative Counts to Daily Rates

### Exercise 5

```{r compute-daily-new-cases}
# Filter to the needed date range
# We restrict to July 5 and August 3.
ks_daily <- ks_cases %>%
  filter(date >= ymd("2020-07-05"), date <= ymd("2020-08-03")) %>%
  group_by(countyFIPS, `County Name`) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(
    daily_new_cases = cumulative_cases - lag(cumulative_cases)
  ) %>%
  ungroup() %>%
  filter(!is.na(daily_new_cases))

glimpse(ks_daily)
```

